---
title: 'GY7702: Coursework 1'
author: "James A. Hardwick (179001497)"
date: "04/11/2020"
output: pdf_document
---
# Introduction 
This document has been created to increase the **reproducibility** of this coursework 
assignment, written in RMarkdown. To support the reproducibility of the document
please refer to the *[GitHub data repository](https://github.com/Jah109/GY7702_CW1.git)*
for the commits that document the development of this Coursework 1

## Libraries 
This coursework use the library **tidyverse**
```{r message = FALSE, warning = FALSE}
library(tidyverse)
```
Also the library **knitr**
```{r message = FALSE, warning = FALSE}
library(knitr)
```

Other libraries are also used for specific question for instance in question 2 the library **palmerpenguins** these specific libraries will be referred to within each question  

**## Questions** 

**### Question 1:** 
Question 1 deals with a vector of 25 numbers between 1 and 7, with each value representing answers to survey questions. Some values are missing. Vector was defined by the question: 

```{r message = FALSE, warning = FALSE}
# vector survey_responses contains 25 elements 
survey_responses <- c(NA, 3, 4, 4, 5, 2, 4, NA, 6, 3, 5, 4, 0, 5, 7, 5, NA, 5, 
                      2, 4, NA, 3, 3, 5, NA)
```

#### Question 1.1:

```{r message = FALSE, warning = FALSE}
# specify the survey_response variable 
survey_responses %>%
  # using na.omit removes NA values from the vector 
  na.omit() %>%
  # is. element will return logical output, if 7 or completely agree was answered in the 
  # survey it will be returned as TRUE 
  is.element(7) %>%
  # is. element will return logical output, if 1 or completely agree was answered in the 
  # survey it will be returned as TRUE
  is.element(1)
```
#### Question 1.2
Extract the indexes who at least somewhat agreed or more (values 5 to 7)

```{r message = FALSE, warning = FALSE}
# Function which see if any value between 5 and 7 and return the index of each 
# survey participant 
which(survey_responses %in% 5:7)
```

**###Question 2:**
Question 2 looks data from Adélie, Chinstrap, and Gentoo penguins observed on islands in the Palmer Archipelago near Palmer Station, Antarctica. Palmerpenguins library can be found at *[Palmerpenguins GitHub Repositry](https://github.com/allisonhorst/palmerpenguins.git)*

#### Question 2.1
Question 2.1 ask for the library (palmerpenguins) to be installed and loaded 
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
# install.packages("palmerpenguins")
library(palmerpenguins)
```

#### Question 2.2
Question 2.2 creates a table showing *species, island, bill length and body mass* of the 10 Gentoo penguins in the penguins table with the highest body mass
```{r message = FALSE, warning = FALSE}
# Starts from the entire palmerpenguins libraries  
palmerpenguins::penguins %>%
  # Selects only the necessary columns
  dplyr::select(species, island, bill_length_mm, body_mass_g
  ) %>%
  # Retain only rows representing the Gentoo species 
  dplyr::filter(species == "Gentoo"
  ) %>%
  # Sort by descending body mass in g
  dplyr::arrange(desc(body_mass_g)) 
```

#### Question 2.3 
Question 2.3 creates a table with *average bill length per island*, ordered by *average bill length*
```{r message = FALSE, warning = FALSE}
# Starts from the entire palmerpenguins libraries  
palmerpenguins::penguins %>%
  # Selects only the necessary columns
  dplyr::select(bill_length_mm, island) %>%
  # Grouped by island
  dplyr::group_by(island) %>%
  # Drops rows containing NAs in the bill_length_mm column
  # otherwise the mean function will return NA
  dplyr::filter(!is.na(bill_length_mm)) %>%
  # Calculates the average of bill_length_mm
  dplyr::summarise(average_bill_length = mean(bill_length_mm)) %>%
  # Ordered by descending average_bill_length
  dplyr::arrange(desc(average_bill_length)) %>%
  # kable improves tibble format 
  knitr::kable()
```

#### Question 2.4 
Question 2.4 creates a table showing the *minimum, median and maximum* proportion between *bill length and bill depth by species* 
```{r message = FALSE, warning = FALSE}
# Starts from the entire palmerpenguins libraries 
palmerpenguins::penguins %>%
  # Selects only the necessary columns
  dplyr::select(species, bill_length_mm, bill_depth_mm) %>%
  # Grouped by species
  dplyr::group_by(species) %>%
  # Drops rows containing NAs in the bill_length_mm column
  # otherwise the mean function will return NA
  dplyr::filter(!is.na(bill_length_mm))%>%
  # Drops rows containing NAs in the bill_depth_mm column
  # otherwise the mean function will return NA
  dplyr::filter(!is.na(bill_depth_mm)) %>%
  # Calculates the bill length to bill depth ratio 
  dplyr::summarise(Proportion= 
                     (bill_length_mm/bill_depth_mm)) %>%
   # using summariase again the minimum, median and maximum for each species can be calculated 
  dplyr::summarise(min(Proportion), 
                   median(Proportion),
                   max(Proportion)) %>%
  # Using the function kable formats the table 
  knitr::kable()
```


**### Question 3:**

Question 3 looks at a topical data set of new and cumulative **COVD19** cases in the UK between March 1st and October 17th 2020. **COVID19** data is sourced from the [HM Government Coronavirus in the UK](https://coronavirus-staging.data.gov.uk/details/cases) 

#### Question 3.1
Question 3.1 asks for the data covid19 cases to be loaded  

```{r message = FALSE, warning = FALSE}
#using readr (part of tidyverse)
library(readr)
# reads the .CSV file with the correct directory
# Imports covid19_cases_20200301_20201017.csv and assings to a new variable
#covid_data 
covid_data <-readr::read_csv("covid19_cases_20200301_20201017.csv")
```

Tyding the data 


#### Question 3.2
This question asked for an area specific table to be generated. The area I was assigned was **Brentwood** in Essex. A summary of   
Brentwood Borough Council *(COVID19 response and information)[https://www.brentwood.gov.uk/index.php?cid=2937]* can be found at this address. During the creation of *Brentwood_complete_covid_data* the original *covid_data* was cleaned using tidyr. 

```{r message = FALSE, warning = FALSE}
# Creates a new table named Brentwood_complete_covid_data from manipulation of 
# imported national covid_data 
Brentwood_complete_covid_data <- covid_data %>%
  # desired columns (specimen_date, area_name, newCasesBySpecimenDate,
  # cumCasesBySpecimenDate) are selected 
  dplyr::select(specimen_date, area_name, newCasesBySpecimenDate,
                cumCasesBySpecimenDate) %>%
  # table then ordered based on specimen_date and area_name 
  dplyr::arrange(specimen_date, area_name) %>%
  # fill replace NA values with the values from the previous row 
  # default direction is down
  tidyr::fill(newCasesBySpecimenDate, cumCasesBySpecimenDate) %>%
  # replaces any remaining NA with 0 
 tidyr::replace_na(list("newCasesBySpecimenDate" = 0,
                         "cumCasesBySpecimenDate"= 0)) %>%
  # filter data to select on values when the area_name equals Bentwood 
  dplyr::filter(area_name == "Brentwood") %>%
  # drops column area_name using select 
  dplyr::select(-area_name)%>%
  # prints a representative number of rows 
  print(n= 8)

```

#### Question 3.3
#### load library lubridate

Initally, library [lubdridate](https://lubridate.tidyverse.org/) is loaded to
aid with date format.
```{r message = FALSE, warning = FALSE}
library("lubridate")
```

Second part of this question converts day_before into character from following 
year month and day format and reformats the the table
```{r message = FALSE, warning = FALSE}
# Creates a new table Brentwood_day_before from manipulation of 
# Brentwood_complete_covid_data 
Brentwood_day_before <- Brentwood_complete_covid_data %>%
  # using lubridate, a new column day_before is created 
  # day_before is in character data type in the format year month day 
  # with specimen_date from Brentwood_complete_covid_data - 1  
  dplyr::mutate(day_before = as.character(ymd(specimen_date -1))) %>%
  # specimen_date and cumCasesBySpecimenDate are droped from the tble 
  dplyr::select(-specimen_date, -cumCasesBySpecimenDate) %>%
  #newCasesBySpecimenDate are renamed to newCases_day_before 
  dplyr::rename(newCases_day_before = newCasesBySpecimenDate) %>% 
  # slice_head to print a representive number of rows 
  dplyr::slice_head(n= 8)
```

Initially, when I tried to join the tables *Brentwood_complete_covid_data* with 
*Brentwood_day_before* this didn't work as each column in the join was in a 
different data type <date> and <Character>. So I tried:

1. converting column specimen_date in *Brentwood_complete_covid_data* resulting in an error when attempted to convert the date to character as specified by the question in the table *Brentwood_day_before* 
2. Making a new table *Brentwood_complete_covid_data_converted* before the table join wassuccessful in enabling the table join. However, it is a messy way round of executing this column data type conversion 2

```{r message = FALSE, warning = FALSE}
# New table Brentwood_complete_covid_data_converted created 
Brentwood_complete_covid_data_converted <- Brentwood_complete_covid_data %>%
  # using mutate with the function across allowed the date value to be 
  # converted to character without the creation of a new column  
  dplyr::mutate(across(where(is.Date), as.character)) 
```

Lastly, Brentwood_day_before and Brentwood_complete_covid_data are joined using
a left join. With day_before being equal to specimen_date both in character data type.
Then a new column is added showing the daily percentage change of cases. 
Stored in a table called *Brentwood_covid_development*

```{r message = FALSE, warning = FALSE}
# Left table_join where specimen_date and day_before are equal 
Brentwood_covid_development <- dplyr::left_join(Brentwood_complete_covid_data_converted, 
                                                Brentwood_day_before, 
                                                by =c("specimen_date" = "day_before")) %>%
  # using mutate new column was created showing the percentage day on day change 
  dplyr::mutate(percentage_new_cases = 
                  ((newCasesBySpecimenDate / newCases_day_before)*100)) %>%
  # Replaces all NaN with 0 
  tidyr::replace_na(list(percentage_new_cases = 0)) %>%
  knitr::kable()

# a number of values in the percentage_new_cases returns Inf, Inf is a product
# of when there are no newCases_day_before 
# will print Brentwood_covid_development, with 20 lines being shown 
print(Brentwood_covid_development, n = 40)
```



#### Question 3.4

495 COVID-19 cases were recorded in the Brentwood area between 2020-03-02 to 2020-10-09. Development of cases was not linear instead three distinct phases can be observed. Phase-1 began between the period 2020-03-03 to 2020-03-16, only two cases where recorded. Between 2020-03-17 to 2020-05-14 cases increased rapidly with percentage of new cases of day before peaking at 800%. During phase-3, new cases began to slow and stabilised overall between 2020-05-15 and 2020-08-19. Daily percentage change fell over time (e.g. 200% to 0%). Within this general trend there was some short-scale variability (e.g. 2020-05-17 to 2020-05-21 of 0% to 25% to 133% before returning to 0%). Stabilisation of cases, during phase-2, was short lived with phase-3 mirroring the initial phase-1 development of new cases starting from 2020-08-20. During Phase-3 percentage of new cases was rapidly rising averaging 100.71% daily increase, a higher magnitude than phase-1 (i.e 84%). 

**###Question 4**

Importing population per local authority and COVID19 case data 
```{r message = FALSE, warning = FALSE}
# import lad19_population data and assigned to a new variable LAD_pop 
LAD_pop<- readr::read_csv("lad19_population.csv")
# import COVID19 case data and assigned to a new variable LAD_covid_cases 
LAD_covid_cases <- readr::read_csv("covid19_cases_20200301_20201017.csv")
```

Carrying out table join for LAD_pop and LAD_covid_cases 
```{r message = FALSE, warning = FALSE}
# need to rename LAD_pop column lad19_area_name to area_name to match
# LAD_covid_cases table 
covid_cases_lad_pop <- dplyr::rename(LAD_pop,area_name = lad19_area_name) %>%
# table join between LAD_pop and LAD_covid_cases 
dplyr::left_join(LAD_covid_cases, LAD_pop, by = "area_name")
```

####Description and Interpretation 

