---
title: 'GY7702: Coursework 1'
author: "James A. Hardwick (179001497)"
date: "04/11/2020"
output: pdf_document
---
# Introduction 
This document has been created to increase the **reproducibility** of this coursework 
assignment, written in RMarkdown. To support the reproducibility of the document
please refer to the [**GitHub data repository**](https://github.com/Jah109/GY7702_CW1.git)
for the commits that document the development of this Coursework 1. 

## Libraries 
This coursework use the library [**tidyverse**](https://www.tidyverse.org/)
```{r message = FALSE, warning = FALSE}
library(tidyverse)
```
Also the library [**knitr**](https://yihui.org/knitr/)
```{r message = FALSE, warning = FALSE}
library(knitr)
```

Other libraries are also used for specific question for instance in question 2 the library [**palmerpenguins**](https://github.com/allisonhorst/palmerpenguins) and in question 3 [**Lubridate**](https://lubridate.tidyverse.org/)

## **Questions** 

### **Question 1:** 
Question 1 deals with a vector of 25 numbers between 1 and 7, with each value representing answers to survey questions. Some values are missing. Vector was defined by the question: 

```{r message = FALSE, warning = FALSE}
# vector survey_responses contains 25 elements 
survey_responses <- c(NA, 3, 4, 4, 5, 2, 4, NA, 6, 3, 5, 4, 0, 5, 7, 5, NA, 5, 
                      2, 4, NA, 3, 3, 5, NA)
```

#### Question 1.1:

```{r message = FALSE, warning = FALSE}
# Starting from the vector Survey_responses
survey_responses %>%
   # first omits NAs from the vector 
  na.omit() %>%
  # using %in% and c(1,7) will check each value for either completely disagrees 
  # or completely agrees
  # as checking for specific values, 0 will be also FALSE  
  (function(x) x %in% c(1,7))


  
```
#### Question 1.2
Extract the indexes who at least somewhat agreed or more (values 5 to 7)

```{r message = FALSE, warning = FALSE}
# Function which see if any value between 5 and 7 and return the index of each 
# survey participant 
which(survey_responses %in% 5:7)
```

### **Question 2:**
Question 2 looks data from Adélie, Chinstrap, and Gentoo penguins observed on islands in the Palmer Archipelago near Palmer Station, Antarctica. Palmerpenguins library can be found at [**Palmerpenguins GitHub Repositry**](https://github.com/allisonhorst/palmerpenguins.git)

#### Question 2.1
Question 2.1 ask for the library (palmerpenguins) to be installed and loaded 
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
# install.packages("palmerpenguins")
library(palmerpenguins)
```

#### Question 2.2
Question 2.2 creates a table showing *species, island, bill length and body mass* of the 10 Gentoo penguins in the penguins table with the highest body mass
```{r message = FALSE, warning = FALSE}
# Starts from the entire palmerpenguins libraries  
palmerpenguins::penguins %>%
  # Selects only the necessary columns
  dplyr::select(species, island, bill_length_mm, body_mass_g
  ) %>%
  # Retain only rows representing the Gentoo species 
  dplyr::filter(species == "Gentoo"
  ) %>%
  # Sort by descending body mass in g
  dplyr::arrange(desc(body_mass_g)) 
```

#### Question 2.3 
Question 2.3 creates a table with *average bill length per island*, ordered by *average bill length*
```{r message = FALSE, warning = FALSE}
# Starts from the entire palmerpenguins libraries  
palmerpenguins::penguins %>%
  # Selects only the necessary columns
  dplyr::select(bill_length_mm, island) %>%
  # Grouped by island
  dplyr::group_by(island) %>%
  # Drops rows containing NAs in the bill_length_mm column
  # otherwise the mean function will return NA
  dplyr::filter(!is.na(bill_length_mm)) %>%
  # Calculates the average of bill_length_mm
  dplyr::summarise(average_bill_length = mean(bill_length_mm)) %>%
  # Ordered by descending average_bill_length
  dplyr::arrange(desc(average_bill_length)) %>%
  # kable improves tibble format 
  knitr::kable()
```

#### Question 2.4 
Question 2.4 creates a table showing the *minimum, median and maximum* proportion between *bill length and bill depth by species* 
```{r message = FALSE, warning = FALSE}
# Starts from the entire palmerpenguins libraries 
palmerpenguins::penguins %>%
  # Selects only the necessary columns
  dplyr::select(species, bill_length_mm, bill_depth_mm) %>%
  # Grouped by species
  dplyr::group_by(species) %>%
  # Drops rows containing NAs in the bill_length_mm column
  # otherwise the mean function will return NA
  dplyr::filter(!is.na(bill_length_mm))%>%
  # Drops rows containing NAs in the bill_depth_mm column
  # otherwise the mean function will return NA
  dplyr::filter(!is.na(bill_depth_mm)) %>%
  # Calculates the bill length to bill depth ratio 
  dplyr::summarise(Proportion= 
                     (bill_length_mm/bill_depth_mm)) %>%
   # using summariase again the minimum, median and maximum for each species can be calculated 
  dplyr::summarise(min(Proportion), 
                   median(Proportion),
                   max(Proportion)) %>%
  # Using the function kable formats the table 
  knitr::kable()
```


### **Question 3:**

Question 3 looks at a topical data set of new and cumulative **COVD19** cases in the UK between March 1st and October 17th 2020. **COVID19** data is sourced from the [**HM Government Coronavirus in the UK**](https://coronavirus-staging.data.gov.uk/details/cases) 

#### Question 3.1

Question 3.1 asks for the data covid19 cases to be loaded  

```{r message = FALSE, warning = FALSE}
#using readr (part of tidyverse)
library(readr)
# reads the .CSV file with the correct directory
# Imports covid19_cases_20200301_20201017.csv and assings to a new variable
#covid_data 
covid_data <-readr::read_csv("covid19_cases_20200301_20201017.csv")
```

#### Question 3.2

This question asked for an area specific table to be generated. The area I was assigned was **Brentwood** in Essex. A summary of Brentwood Borough Council (**COVID19 response and information**) [https://www.brentwood.gov.uk/index.php?cid=2937] can be found at this address. During the creation of *Brentwood_complete_covid_data* the original *covid_data* was cleaned using tidyr. 

```{r message = FALSE, warning = FALSE}
# Creates a new table named Brentwood_complete_covid_data from manipulation of 
# imported national covid_data 
Brentwood_complete_covid_data <- covid_data %>%
  # desired columns (specimen_date, area_name, newCasesBySpecimenDate,
  # cumCasesBySpecimenDate) are selected 
  dplyr::select(specimen_date, area_name, newCasesBySpecimenDate,
                cumCasesBySpecimenDate) %>%
  # table then ordered based on specimen_date and area_name 
  dplyr::arrange(specimen_date, area_name) %>%
  # fill replace NA values with the values from the previous row 
  # default direction is down
  tidyr::fill(newCasesBySpecimenDate, cumCasesBySpecimenDate) %>%
  # replaces any remaining NA with 0 
 tidyr::replace_na(list("newCasesBySpecimenDate" = 0,
                         "cumCasesBySpecimenDate"= 0)) %>%
  # filter data to select on values when the area_name equals Bentwood 
  dplyr::filter(area_name == "Brentwood") %>%
  # drops column area_name using select 
  dplyr::select(-area_name)%>%
  # prints a representative number of rows 
  print(n= 8)

```

#### Question 3.3

Initially, library (**lubdridate**)[https://lubridate.tidyverse.org/] is loaded to
aid with date format.

```{r message = FALSE, warning = FALSE}
library("lubridate")
```

Second part of this question converts day_to_match into character from following 
year month and day format and reformat the table.
```{r message = FALSE, warning = FALSE}
# Creates a new table Brentwood_day_before from manipulation of 
# Brentwood_complete_covid_data 
Brentwood_day_before <- Brentwood_complete_covid_data %>%
  # using lubridate, a new column day_to_match is created 
  # day_to_match is in character data type in the format year month day 
  # with specimen_date from Brentwood_complete_covid_data + 1 
  # altered after email from Stef 
  dplyr::mutate(day_to_match = as.character(ymd(specimen_date + 1))) %>%
  # specimen_date and cumCasesBySpecimenDate are dropped from the table 
  dplyr::select(-specimen_date, -cumCasesBySpecimenDate) %>%
  #newCasesBySpecimenDate are renamed to newCases_day_before 
  dplyr::rename(newCases_day_before = newCasesBySpecimenDate)
  
print(Brentwood_day_before)
  
```

Initially, when I tried to join the tables *Brentwood_complete_covid_data* with 
*Brentwood_day_before* this didn't work as each column in the join was in a 
different data type <date> and <Character>. So I tried:

1. converting column specimen_date in *Brentwood_complete_covid_data* resulting in an error when attempted to convert the date to character as specified by the question in the table *Brentwood_day_before* 
2. Making a new table *Brentwood_complete_covid_data_converted* before the table join was successful in enabling the table join. However, it is a messy way round of executing this column data type conversion 

```{r message = FALSE, warning = FALSE}
# New table Brentwood_complete_covid_data_converted created 
Brentwood_complete_covid_data_converted <- Brentwood_complete_covid_data %>%
  # using mutate with the function across allowed the date value to be 
  # converted to character without the creation of a new column  
  dplyr::mutate(across(where(is.Date), as.character)) 
```

Lastly, Brentwood_day_before and Brentwood_complete_covid_data are joined using
a left join. With day_to_match being equal to specimen_date both in character data type.
Then a new column is added showing the daily percentage change of cases. 
Stored in a table called *Brentwood_covid_development*

```{r message = FALSE, warning = FALSE}
# Left table_join where specimen_date and day_to_match are equal 
Brentwood_covid_development <- dplyr::left_join(Brentwood_day_before,Brentwood_complete_covid_data_converted, 
                                                by =c("day_to_match"="specimen_date")) %>%
  # using mutate new column was created showing the percentage day on day change 
  dplyr::mutate(per_new_cases = 
                  ((newCasesBySpecimenDate / newCases_day_before)*100)) %>% 
  # Replaces all NaN with 0 
  tidyr::replace_na(list(per_new_cases = 0)) %>%
  # prints table 

  print(as_tibble(Brentwood_covid_development), n = 20)

```



#### Question 3.4

495 COVID-19 cases were recorded in the Brentwood area between 2020-03-02 to 2020-10-09. Development of cases was not linear instead three distinct phases can be observed. Phase-1 began between the period 2020-03-03 to 2020-03-16, only two cases where recorded. Between 2020-03-17 to 2020-05-14 cases increased rapidly with percentage of new cases of day before peaking at 800%. During phase-3, new cases began to slow and stabilised overall between 2020-05-15 and 2020-08-19. Daily percentage change fell over time (e.g. 200% to 0%). Within this general trend there was some short-scale variability (e.g. 2020-05-17 to 2020-05-21 of 0% to 25% to 133% before returning to 0%). Stabilisation of cases, during phase-2, was short lived with phase-3 mirroring the initial phase-1 development of new cases starting from 2020-08-20. During Phase-3 percentage of new cases was rapidly rising averaging 100.71% daily increase, a higher magnitude than phase-1 (i.e 84%). 

### **Question 4**

Importing LAD data sets: 
```{r message = FALSE, warning = FALSE}
# import lad19_population data and assigned to a new variable LAD_pop 
LAD_pop<- readr::read_csv("lad19_population.csv")
# import COVID19 case data and assigned to a new variable LAD_covid_cases 
LAD_covid_cases <- readr::read_csv("covid19_cases_20200301_20201017.csv")
```

**Aim of Analysis**

* To examine if there is a north south divide in England of the development of COVID19 cases over time

**Objectives**

* To select a northern local authority district with similar population (~77,032) to Brentwood (my assigned area) 
* Compare the magnitude of change in COVID19 Cases 

```{r message = FALSE, warning = FALSE}
# Created LAD_selection 
LAD_selection <- LAD_pop %>%
  # arranged by population and area code enabled a local authority district to be 
  # selected of similar size to Brentwood 
  dplyr::arrange(area_population)
```


[**Rossendale Borough Council**](https://www.rossendale.gov.uk/) was selected as it had a similar population 
(71,525) and located in Lancashire. 

```{r message = FALSE, warning = FALSE}
# Creates new a new tibble from imported data 
Rossendale_covid_data <- covid_data %>%
  # selects desired columns 
  dplyr::select(specimen_date, area_name, newCasesBySpecimenDate,
                cumCasesBySpecimenDate) %>%
  # arrange rows 
  dplyr::arrange(specimen_date, area_name) %>%
  # fill down all NAs
  tidyr::fill("newCasesBySpecimenDate", "cumCasesBySpecimenDate") %>%
  # when unable to fill NA, NA will be equal to 0 
  tidyr::replace_na(list("newCasesBySpecimenDate" = 0,
                         "cumCasesBySpecimenDate"= 0)) %>%
  # filter so that only Rossendale is present in the tibble 
  dplyr::filter(area_name == "Rossendale") %>%
  # drops the area_name column 
  dplyr::select(-area_name) %>%
  # rename column in order to identify columns in the fininal tibble 
  dplyr::rename("R_Cumulative_cases" = "cumCasesBySpecimenDate", 
                "R_new_cases" = "newCasesBySpecimenDate")
```

**Repeat above step for Brentwood**
```{r message = FALSE, warning = FALSE}
# Creates new a new tibble from imported data 
Brentwood_covid_data <- covid_data %>%
  # selects desired columns 
  dplyr::select(specimen_date, area_name, newCasesBySpecimenDate,
                cumCasesBySpecimenDate) %>%
  # arrange rows
  dplyr::arrange(specimen_date, area_name) %>%
  # fill down all NAs
  tidyr::fill("newCasesBySpecimenDate", "cumCasesBySpecimenDate") %>%
  # when unable to fill NA, NA will be equal to 0 
  tidyr::replace_na(list("newCasesBySpecimenDate" = 0,
                         "cumCasesBySpecimenDate"= 0)) %>%
  dplyr::filter(area_name == "Brentwood") %>%
  # drops the area_name column 
  dplyr::select(-area_name) %>%
   # rename column in order to identify columns in the fininal tibble 
  dplyr::rename("B_Cumulative_cases" = "cumCasesBySpecimenDate", 
                "B_new_cases" = "newCasesBySpecimenDate" )

```


#### Description and Interpretation 

Two English local authority districts, Brentwood (Essex) and Rossendale (Lancashire) were compared. To assess the north-south difference in the development of COVID19 cases. During the period 2020-03-03 to 2020-10-10. Brentwood and Rossendale were selected as both have a similar population, 77,032 and 71,525 respectively. Comparison of daily cumulative percentage difference and reported cases was used to explore the magnitude of change between Brentwood and Rossendale. 

At end of 2020-10-10, Rossendale recorded 1105 cases in stark contrast to Brentwood recording just 499 cases. Despite Rossendale, not recording the first case until 2020-03-20 seventeen days before Brentwood (8 cases). Daily percentage difference between Brentwood and Rossendale steadily decreases becoming even (0.00 %) on 2020-05-17. Before cases reported in Rossendale, slow at first, begin to accelerate reaching 75% on 2020-10-10. Reflected in average cases per week in early October, 43 and 8 in Rossendale and Brentwood respectively. 

 A north-south divided is demonstrated by the comparison of cumulative cases and daily percentage difference, of Brentwood and Rossendale. Driving this regional disparity is ‘southern/ London centric’ government view of easing the first national lockdown, in early July 2020. Development of cases in Brentwood continued to slow (33% between 2020-05-01 to 2020-07-04) in contrast Rossendale had further proliferation of cases (40% between 2020-05-01 to 2020-07-04). Differences in COVID19 case development over time strongly suggest that national lockdown restrictions were lift too early. Other factors may also contribute to the north-south 

**Final Table for Description and Interpretation**

```{r message = FALSE, warning = FALSE}
# Create new table named Rossendale_Brentwood_Comparison
# using a full_join, tables joined by specimen date 
Rossendale_Brentwood_Comparison <- dplyr::full_join(Rossendale_covid_data, Brentwood_covid_data) %>%
  # select desired columns 
  dplyr::select(specimen_date, R_Cumulative_cases, R_new_cases,B_Cumulative_cases,B_new_cases) %>%
  # arranges rows by specimen date 
  dplyr::arrange(specimen_date) %>%
  # rename specimen_date date
  dplyr::rename(Date = specimen_date)%>%
  # creates new column to show daily percentage difference 
  dplyr::mutate(per_daily_diff = (R_Cumulative_cases - B_Cumulative_cases)/
                  ((R_Cumulative_cases +B_Cumulative_cases)/2)*100) %>%
  # replace NA with 0 
   tidyr::replace_na(list("B_Cumulative_cases" = 0,
                         "R_Cumulative_cases"= 0,
                         "per_daily_diff" = 0,
                         "R_new_cases" = 0,
                         "B_new_cases" = 0)) %>%
  dplyr::slice_head(n = 20) %>%
  # improve table formatt 
  kable()

print(Rossendale_Brentwood_Comparison)
```


















