---
title: 'GY7702: Coursework 1'
author: "James A. Hardwick (179001497)"
date: "04/11/2020"
output: pdf_document
---
# Introduction 
This document has been created to increase the **reproducibility** of this coursework 
assignment, written in RMarkdown. To support the reproducibility of the document
please refer to the [GitHub data repository](https://github.com/Jah109/GY7702_CW1.git) 
for the commits that document the development of this Coursework 1

## Libraries 
This coursework use the library **tidyverse**
```{r message = FALSE, warning = FALSE}
library(tidyverse)
```
Also the library **knitr**
```{r message = FALSE, warning = FALSE}
library(knitr)
```

Other libraries are also used for specific question for instance in question 2 the library **palmerpenguins** these specific libraries will be referred to within each question  

## Questions 
### Question 1: 
Question 1 deals with a vector of 25 numbers between 1 and 7, with each value representing answers to survey questions. Some values are missing. 
#### Question 1.1:


### Question 2:
Question 2 looks data from Adélie, Chinstrap, and Gentoo penguins observed on islands in the Palmer Archipelago near Palmer Station, Antarctica. Palmerpenguins library can be found at [Palmerpenguins GitHub Repositry](https://github.com/allisonhorst/palmerpenguins.git)

#### Question 2.1
Question 2.1 ask for the library (palmerpenguins) to be installed and loaded 
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
# install.packages("palmerpenguins")
library(palmerpenguins)
```

#### Question 2.2
Question 2.2 creates a table showing *species, island, bill length and body mass* of the 10 Gentoo penguins in the penguins table with the highest body mass
```{r message = FALSE, warning = FALSE}
# Starts from the entire palmerpenguins libraries  
palmerpenguins::penguins %>%
  # Selects only the necessary columns
  dplyr::select(species, island, bill_length_mm, body_mass_g
  ) %>%
  # Retain only rows representing the Gentoo species 
  dplyr::filter(species == "Gentoo"
  ) %>%
  # Sort by descending body mass in g
  dplyr::arrange(desc(body_mass_g)) 
```

#### Question 2.3 
Question 2.3 creates a table with *average bill length per island*, ordered by *average bill length*
```{r message = FALSE, warning = FALSE}
# Starts from the entire palmerpenguins libraries  
palmerpenguins::penguins %>%
  # Selects only the necessary columns
  dplyr::select(bill_length_mm, island) %>%
  # Grouped by island
  dplyr::group_by(island) %>%
  # Drops rows containing NAs in the bill_length_mm column
  # otherwise the mean function will return NA
  dplyr::filter(!is.na(bill_length_mm)) %>%
  # Calculates the average of bill_length_mm
  dplyr::summarise(average_bill_length = mean(bill_length_mm)) %>%
  # Ordered by descending average_bill_length
  dplyr::arrange(desc(average_bill_length)) %>%
  # kable improves tibble format 
  knitr::kable()
```

#### Question 2.4 
Question 2.4 creates a table showing the *minimum, median and maximum proportion* between *bill length and bill depth by species* 
```{r message = FALSE, warning = FALSE}
# Starts from the entire palmerpenguins libraries 
palmerpenguins::penguins %>%
  # Selects only the necessary columns
  dplyr::select(species, bill_length_mm, bill_depth_mm) %>%
  # Grouped by species
  dplyr::group_by(species) %>%
  # Drops rows containing NAs in the bill_length_mm column
  # otherwise the mean function will return NA
  dplyr::filter(!is.na(bill_length_mm))%>%
  # Drops rows containing NAs in the bill_depth_mm column
  # otherwise the mean function will return NA
  dplyr::filter(!is.na(bill_depth_mm)) %>%
  # Calculates the bill length to bill depth ratio 
  dplyr::summarise(bill_length_to_bill_depth_ratio = 
                     (bill_length_mm/bill_depth_mm))
```
